---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data
```{r preprocessing}
# Load the Activity data
activity <- read.csv("activity.csv", header = TRUE, stringsAsFactors = TRUE, colClasses = c("numeric","character","numeric")) 
# Convert to a Date
activity$date <- as.Date(activity$date, format = "%Y-%m-%d", tz = "UTC")
```

## What is mean total number of steps taken per day?
```{r TotalStepsHistogram}
# Aggregate the Total steps by Day
activity.total.byDay <- aggregate(steps ~ date, FUN = sum, data = activity)
# Histogram of total daily steps
hist(activity.total.byDay$steps, main = "Total Daily Steps", xlab = "Total Steps")
# Mean & Median total number of steps taken per day
summary(activity.total.byDay$steps)
```
The Mean Steps is 10766
The Mediam Steps is 10765

## What is the average daily activity pattern?
```{r AverageDailyPatterm}
# Aggregate the Interval Mean Steps
activity.daily.pattern <- aggregate(steps ~ interval, FUN = mean, data = activity)
# Plot the Intervals with Mean Steps
plot(activity.daily.pattern
     , main = "Daily Activity Pattern"
     , xlab = "5 min Interval"
     , ylab = "Mean Steps"
     , type = "l"
     )
# The time interval with the Max Mean Daily Steps
activity.daily.pattern[activity.daily.pattern$steps == max(activity.daily.pattern$steps), ]
```
## Imputing missing values
```{r ImputeMissingValues}
# Number of NA cases in the data
sum(is.na(activity$steps))
# Fill in the NA values with the Mean for that time interval from the overall dataset
activity.na.to.mean <- merge(x = activity[is.na(activity$steps),] 
      ,y = activity.daily.pattern 
      ,by = "interval"
      )
# Copy the original dataset to hold the dataset with imputed values
activity.imputed <- activity
# Change the NA values to the imputed values
activity.imputed[is.na(activity.imputed$steps), "steps"] <- activity.na.to.mean$steps.y
```
Show the histogram for the imputed dataset
```{r ImputedHistogram}
# Aggregate the Total steps by Day
activity.imputed.total.byDay <- aggregate(steps ~ date, FUN = sum, data = activity.imputed)
# Histogram of total daily steps
hist(activity.imputed.total.byDay$steps, main = "Total Daily Steps (Imputed NAs to Mean)", xlab = "Total Steps")
```
Compare the summaries of the Original and Imputed datasets
```{r}
# Imputed - Mean & Median total number of steps taken per day
summary(activity.imputed.total.byDay$steps)
# Original - Mean & Median total number of steps taken per day
summary(activity.total.byDay$steps)
# Difference between the Original and Imputed Means
mean((activity.imputed.total.byDay$steps)) - mean((activity.total.byDay$steps))
```
Original
The Mean Steps is 10766
The Mediam Steps is 10765

Imputed
The Mean Steps is 10766
The Mediam Steps is 11015

There is no difference in the Means between the Original and Imputed dataset
The Mean is a little higher on the Imputed dataset
```{r}
# Difference between the Original and Imputed Totals of the days
sum((activity.imputed.total.byDay$steps)) - sum((activity.total.byDay$steps))
```
There is an increase of 86,129,51 in Total steps in the Imputed over the Originalß dataset

## Are there differences in activity patterns between weekdays and weekends?
```{r weekday}
# Flag Weekday v Weekend
activity.imputed$DOW <- ifelse( 
    weekdays(activity.imputed$date, abbreviate = TRUE) %in% c("Sat", "Sun"), "Weekend"
    , "Weekday")
# Aggregate the Mean steps
activity.imputed.DOW <- aggregate(steps ~ DOW + interval, FUN = mean, data = activity.imputed)
# Plot the Intervals with Mean Steps
library(ggplot2)
qplot(interval, steps
      , data = activity.imputed.DOW
      , facets = DOW ~ .
      , main = "Weekday v Weekend Activity Pattern"
      , xlab = "5 min Interval"
      , ylab = "Mean Steps"
      , geom = "line"
      )
```

